Learning words by recursive pragmatic reasoning about other agents: The blessing of social anxiety

### 2013-05-31 praglearn.tex

\documentclass{article} % For LaTeX2e
\usepackage{nips13submit_e,times}
\usepackage{hyperref}
\usepackage{url}
\usepackage{amsmath}
\usepackage{graphicx}
\usepackage{booktabs}
\usepackage{subcaption}

@@

\begin{table}[t]
\label{tab:results}
\begin{center}
\begin{tabular}{lccccc}
\toprule
Phenomenon & Ref. & WL & PI & PI+U & This work \\
\midrule
Interpreting scalar implicature & \cite{grice1975} &  & x &x & x\\
Interpreting Horn implicature & \cite{horn1984} & &  & x& x\\
Cross-situational learning & \cite{smith2008} & x& & & x \\
Disambiguating new words using old words & \cite{markman1988} &x  & & & x \\
Emergence of novel \& efficient communicative systems & \cite{galantucci2005} & & & & x \\
Learning scalar implicature & & & & & x \\
\bottomrule
\end{tabular}
\end{center}
\caption{Empirical results and references. WL refers to the word learning model of \cite{frank2009}; PI refers to the recursive pragmatic inference model of \cite{frank2012}; PI+U refers to the pragmatic inference model of \cite{bergen2012} which includes lexical uncertainty, marginalizes it out, and then recurses.}
\end{table}

\subsection{Scalar implicature}

Many sets of words in natural language form scales in which each term refers to successively stronger restrictions on the situations compatible with the term. ``Some'' and ``all'' form a scale of this type. While ``I ate some of the cookies'' is compatible with the followup ``in fact, I ate {\it all} of the cookies,'' the reverse is not true. ``Might'' and ``must'' are another example, as are ``OK,'' ``good,'' and ``excellent.'' All of these scales allow for {\it scalar implicatures} \cite{grice1975}: if a speaker uses a weaker term, a listener is licensed to infer that the stronger term was not true in this situation. So although ``I ate some of the cookies'' could in principle be compatible with eating all of them, the listener is lead to believe that ``some but not all'' is the likely state of affairs.

The recursive pragmatic reasoning portions of our model capture set of findings on scalar implicature effectively. Consider a lexicon
$\left[
    \begin{array}{cc}
      0.5 & 0.5 \\
      0 & 1
    \end{array} 
  \right]$,
where rows show the words ``some'' and ``all'' and columns indicate worlds where {\sc some but not all} and {\sc all} are true, respectively, capturing the regularity that ``some'' can be said of situations where {\sc all} is actually the case. Assuming uniform costs for the two words, a rational speaker confronted with such a lexicon will prefer to say ``some'' only when they mean \textsc{some but not all}, and this is therefore the interpretation that a rational listener will place on ``some'' as well. To simulate this situation, we place a Dirichlet(10, 10) prior on the meaning of \textit{some} and a Dirichlet(1, 10) prior on the meaning of \textit{all}, and find this gives speaker use probabilities of 
$\left[
    \begin{array}{cc}
      1.00 & 0.00 \\
      0.00 & 1.00
    \end{array} 
  \right]$
and listener interpretation probabilities of 
$\left[
    \begin{array}{cc}
      .86 & .14 \\
      .03 & .97
    \end{array} 
  \right]$.

@@

Humans, however, unlike traditional game theoretic models, make the inference that if two utterances seem equivalent except that one is more costly to utter, then the costly utterance refers to an unusual event. We call this pattern \textit{Horn implicature}, after \cite{horn1984}. For instance, ``Lee got the car to stop'' implies that Lee used an unusual method (e.g.~not the brakes) because, had he used the brakes, the speaker would have chosen the simpler and shorter (less costly) expression, ``Lee stopped the car'' \cite{horn1984}. Surprisingly, Bergen et al \cite{bergen2012} show that the key to achieving this favorable result is ignorance. If a listener assigns equal probability to her partner using the good system or the bad system, then their best bet is to estimate $P_S(\word | \obj)$ as the average of $P_S(\word | \obj, \text{good system})$ and $P_S(\word | \obj, \text{bad system})$. These might seem to cancel out, but in fact they do not. In the good system, the speaker has higher expected utility than the bad system; therefore, a soft-max agent in the bad system has noiser behaviour than in the good system, and their behaviour in the good system dominates the average, breaking the symmetry. Similar reasoning applies to an uncertain speaker.

@@

The model of \cite{bergen2012} differs from ours in that they then apply further layers of pragmatic recursion on top of these marginal distributions, which as described above removes the possibility of learning; in their model it is only hypothetical agents who are ignorant, while the actual speaker and listener know everything there is to know about each other's behaviour. However, the core result carries over to our model. For example, with a uniform prior over lexicons and $P_\text{prior}(\text{\textsc{common}}) = 0.8$, $\text{cost}(\text{``cheap''}) = 0.5$, $\text{cost}(\text{``expensive''}) = 1.0$, the symmetry breaks in the appropriate way: Despite total ignorance of the conventional system, speakers prefer to use simple words for common referents ($P_S(\text{``cheap''} | \text{\textsc{common}}) = 0.88, P_S(\text{``cheap''} | \text{\textsc{rare}}) = 0.46$), and listeners show a similar bias ($P_L(\text{\textsc{common}} | \text{``cheap''}) = 0.77, P_L(\text{\textsc{common}} | \text{``expensive''}) = 0.65$).

@@

A recent line of work in language acquisition has 
% \begin{figure}[t]
%   \centering
%   \includegraphics[width=0.20\textwidth]{figures/cross-sit.pdf}
%   \caption{Cross-situational simulation}
%   \label{fig:cross-sit}
% \end{figure}

>>>>

\documentclass{article} % For LaTeX2e
\usepackage{nips13submit_e,times}
\usepackage{hyperref}
\usepackage{url}
\usepackage{amsmath}
\usepackage{graphicx}
\usepackage{booktabs}
\usepackage[raggedright]{sidecap}

@@

\begin{table}[t]
\label{tab:results}
\begin{center}
\begin{tabular}{lccccc}
\toprule
Phenomenon & Ref. & WL & PI & PI+U & This work \\
\midrule
Interpreting specificity implicature & \cite{grice1975} &  & x &x & x\\
Interpreting Horn implicature & \cite{horn1984} & &  & x& x\\
Cross-situational learning & \cite{smith2008} & x& & & x \\
Disambiguating new words using old words & \cite{markman1988} &x  & & & x \\
Disambiguation without mapping & \cite{horst2008} &  & & & x \\
Emergence of novel \& efficient communicative systems & \cite{galantucci2005} & & & & x \\
Lexicalization of Horn implicature & \cite{horn1984} & & & & x \\
Lexicalization of specificity implicature & \cite{levinson2000} & & & & x \\
\hline
\end{tabular}
\end{center}
\caption{Empirical results and references. WL refers to the word learning model of \cite{frank2009}; PI refers to the recursive pragmatic inference model of \cite{frank2012}; PI+U refers to the pragmatic inference model of \cite{bergen2012} which includes lexical uncertainty, marginalizes it out, and then recurses.}
\end{table}

\subsection{Specificity implicature}

Many sets of words in natural language form scales in which each term refers to successively stronger restrictions on the situations compatible with the term. ``Some'' and ``all'' form a scale of this type. While ``I ate some of the cookies'' is compatible with the followup ``in fact, I ate {\it all} of the cookies,'' the reverse is not true. ``Might'' and ``must'' are another example, as are ``OK,'' ``good,'' and ``excellent.'' All of these scales allow for {\it scalar implicatures} \cite{grice1975}: if a speaker uses a weaker term, a listener is licensed to infer that the stronger term was not true in this situation. So although ``I ate some of the cookies'' could in principle be compatible with eating all of them, the listener is lead to believe that ``some but not all'' is the likely state of affairs. We use the term {\it specificity implicature} here (following \cite{bergen2012}) to denote that we are agnostic about whether such implicatures are conventionalized or whether they use a contextually-defined scale.

The recursive pragmatic reasoning portions of our model capture findings on specificity implicature. Consider a representation of scalar implicature as a matrix 
$\left[
    \begin{array}{cc}
      0.5 & 0.5 \\
      0 & 1
    \end{array} 
  \right]$,
where rows show the words ``some'' and ``all'' and columns indicate worlds where {\sc some but not all} and {\sc all} are true, respectively, capturing the regularity that ``some'' can be said of situations where {\sc all} is actually the case. Assuming uniform costs for the two words, a rational speaker confronted with such a lexicon will prefer to say ``some'' only when they mean \textsc{some but not all}, and this is therefore the interpretation that a rational listener will place on ``some'' as well. To simulate this situation, we place a Dirichlet(10, 10) prior on the meaning of \textit{some} and a Dirichlet(1, 10) prior on the meaning of \textit{all}, and find this gives speaker use probabilities of 
$\left[
    \begin{array}{cc}
      1.00 & 0.00 \\
      0.00 & 1.00
    \end{array} 
  \right]$
and listener interpretation probabilities of 
$\left[
    \begin{array}{cc}
      .86 & .14 \\
      .03 & .97
    \end{array} 
  \right]$.

@@

Humans, however, unlike traditional game theoretic models, do make the inference that if two utterances seem equivalent except that one is more costly to utter, then the costly utterance refers to an unusual event. We call this pattern \textit{Horn implicature}, after \cite{horn1984}. For instance, ``Lee got the car to stop'' implies that Lee used an unusual method (e.g.~not the brakes) because, had he used the brakes, the speaker would have chosen the simpler and shorter (less costly) expression, ``Lee stopped the car'' \cite{horn1984}. Surprisingly, Bergen et al \cite{bergen2012} show that the key to achieving this favorable result is ignorance. If a listener assigns equal probability to her partner using the good system or the bad system, then their best bet is to estimate $P_S(\word | \obj)$ as the average of $P_S(\word | \obj, \text{good system})$ and $P_S(\word | \obj, \text{bad system})$. These might seem to cancel out, but in fact they do not. In the good system, the speaker has higher expected utility than the bad system; therefore, a soft-max agent in the bad system has noiser behaviour than in the good system, and their behaviour in the good system dominates the average, breaking the symmetry. Similar reasoning applies to an uncertain speaker.

% Mike's text:
% Such an implicature again arises naturally from our model. We create a situation in which expressions have differential costs (e.g. $[.5 1.0]$) and referents have differential probabilities (e.g. $[.8 .2]$). At the settings used throughout the manuscript, speakers in such a situation choose simple words to refer to common referents at different rates than for uncommon references. This situation breaks the symmetry though does not result in as dramatic a diagonal as in the specificity implicature, e.g. for the speaker, \footnote{Rows indicate words and columns indicate referents. For speakers, columns are normalized to indicate a distribution over words to be produced given referents, and for listeners rows are normalized to indicate a distribution on referents given words heard.}
% $\left[
%     \begin{array}{cc}
%       .88 & .54 \\
%       .12 & .46
%     \end{array} 
%   \right]$ 
% and for the listener,
% $\left[
%     \begin{array}{cc}
%       .77 & .23 \\
%       .65 & .35
%     \end{array} 
%   \right]$. See Section \ref{sec:hornlearn} for the consequences of repeating this implicature until it is lexicalized.

@@

The model of \cite{bergen2012} differs from ours in that they then apply further layers of pragmatic recursion on top of these marginal distributions, which as described above removes the possibility of learning; in their model it is only hypothetical agents who are ignorant, while the actual speaker and listener know everything there is to know about each other's behaviour. However, the core result carries over to our model. For example, with a uniform prior over lexicons and $P_\text{prior}(\text{\textsc{common}}) = 0.8$, $\text{cost}(\text{``cheap''}) = 0.5$, $\text{cost}(\text{``expensive''}) = 1.0$, the symmetry breaks in the appropriate way: Despite total ignorance about the conventional system, speakers prefer to use simple words for common referents ($P_S(\text{``cheap''} | \text{\textsc{common}}) = 0.88, P_S(\text{``cheap''} | \text{\textsc{rare}}) = 0.46$), and listeners show a similar bias ($P_L(\text{\textsc{common}} | \text{``cheap''}) = 0.77, P_L(\text{\textsc{common}} | \text{``expensive''}) = 0.65$). This preference is somewhat weak, at least to start -- see \S\ref{sec:horn-emergence}

@@

Even very young children can learn words from individually ambiguous situations, e.g. where multiple words are spoken and multiple objects are present \cite{smith2008}. While no individual situation gives away the mapping, the pattern of co-occurrences between words and objects allows for learning. Our current model reproduces this pattern of results, converging to correct lexical mappings from ambiguous evidence. A representative simulation is shown in Fig. \ref{fig:cross-sit}.

\begin{SCfigure}[2.5]
  \includegraphics[width=0.2\textwidth]{figures/cross-sit-mod.pdf}
  \caption{Results from a simulation of cross-situational word learning. The posterior distribution over the learner's lexicon (diagonal entries are correct for word/object pairs 1 -- 3; word D is a distractor). Training data were situations including two of the three objects and a single word, either the distractor or a word matching one of the two objects. Crucially, at no point does the learner observe the meaning of any word; they observe only that a word was produced in each situation.}
  \label{fig:cross-sit}
\end{SCfigure}
